---
title: "SignallingSingleCell"
geometry: margin=0.5cm
author: "Kyle Gellatly"
date: "`r Sys.Date()`"
output:
html_document:
  highlight: tango
  theme: cerulean
  toc: yes
  toc_float : yes
# pdf_document:
#   highlight: tango
#   theme: cerulean
#   toc: yes
#   toc_float : yes
# output: rmarkdown::html_vignette
# vignette: >
#   %\VignetteIndexEntry{Vignette Title}
#   %\VignetteEngine{knitr::rmarkdown}
#   %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE, warning=FALSE, error=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "")
```


```{r, include=FALSE, cache=FALSE, warning=FALSE, error=FALSE}
library("SignallingSingleCell")
load("/Users/KGellatly/Dropbox (UMass Medical School)/Lab_inDrop/mDC_vignette_data/mDC_0hr_1hr_4hr_CLEAN.Rdata") # this in in the Lab_inDrop folder so anyone can access it!!
sc_dat <- mDC_0hr_1hr_4hr_CLEAN
```

# Preprocessing  

## Constructing the ExpressionSet Class  

The ExpressionSet class (ex_sc) is an extremely convienient data structure that contains 3 dataframes. These dataframes contain expression data, cell information, and gene information respectivelty. 

exprs(ex_sc) is the expression data  
pData(ex_sc) is cell information, where rows are cells and columns are metadata  
fData(ex_sc) is gene information, where rows are genes and columns are metadata  
ncol(exprs(ex_sc)) == nrow(pData(ex_sc))  
nrow(exprs(ex_sc)) == nrow(fData(ex_sc))  

```{r, include=TRUE, cache=FALSE, warning=FALSE, error=FALSE}
ex_sc_example <- construct_ex_sc(sc_dat) # sc_dat == Input expression matrix
ex_sc_example # Note that phenoData and featureData are empty right now!
```

```{r, include=FALSE, cache=FALSE, warning=FALSE, error=FALSE}
rm(mDC_0hr_1hr_4hr_CLEAN, sc_dat)
```

## Filtering

The first step is to filter your data to remove low quality cells. Often creating a histogram of the values and assigning cutoffs is simple and effective. Typically we remove all cells lower than 500-1000 UMIs / cell.

```{r, include=TRUE, cache=FALSE, warning=FALSE, error=FALSE}
ex_sc_example <- calc_libsize(ex_sc_example) # sums counts for each cell
plot_density(ex_sc_example, val = "UMI_sum", statistic = "mean") # Density function. Note how many low UMI count cells are in the data!

ex_sc_example <- pre_filter(ex_sc_example, threshold = 1, minCells = 10, minUMI = 1000, maxUMI = 10000) # filters cells and genes
ex_sc_example <- calc_libsize(ex_sc_example)
plot_density(ex_sc_example, val = "UMI_sum", statistic = "mean")  

# save_ggplot(filename = "Density", format = "pdf") # Plots can be quickly and easily saved!
```

## Dimension reduction

Before normalization dimensionality reduction is necessary to form preliminary clusters. These clusters are used to normalize internal to a cluster before normalizing across clusters. First we can subset the genes, and then use these feature selected genes for dimension reduction.

```{r, include=TRUE, cache=FALSE, warning=FALSE, error=FALSE}
gene_subset <- subset_genes(ex_sc_example, method = "PCA", threshold = 1, minCells = 30, nComp = 10, cutoff = 0.85) # 3 available methods for selecting genes

ex_sc_example <- dim_reduce(ex_sc_example, genelist = gene_subset, pre_reduce = "iPCA", nComp = 15, tSNE_perp = 30, iterations = 500, print_progress=TRUE)  # 3 available methods for dimensionality reduction

colnames(pData(ex_sc_example))

plot_tsne_metadata(ex_sc_example, color_by = "UMI_sum", title = "Total UMIs per cell") 

```

## Initial Clustering

Now that we have dimension reduced data we can try clustering it!

```{r, include=TRUE, cache=FALSE, warning=FALSE, error=FALSE}
ex_sc_example <- cluster_sc(ex_sc_example, dimension = "Comp", method = "spectral", num_clust = 6) # 2 clustering algorithms can be performed on 2d plots or PCA / ICA components

plot_tsne_metadata(ex_sc_example, color_by = "Cluster", title = "Spectral Cluster on iPCA components") # this function is flexible for plotting any data onto a 2d tsne mapping

plot_density(ex_sc_example, val = "UMI_sum", color_by = "Cluster", statistic = "mean")
```

## Normalization

Now that the data has preliminary clusters, we can normalize. CPM and TMM normalization will not utilize these clusters, however the much more powerful SCRAN normalization will first normalize internally in clusters, before normalizing acromarker_facets clusters. Once the data is normalized we can run the same steps as above before visualization.

```{r, include=TRUE, cache=FALSE, warning=FALSE, error=FALSE}
```

# Basic Analysis

## Data Annotation

Often we have metadata about the experiment that can be valuable in the analysis! Writing that information now may be appropriate. Our experiment consists of a time course with LPS stimulation. Now we can begin to take advantage of our faceting!

```{r, include=TRUE, cache=FALSE, warning=FALSE, error=FALSE}
pData(ex_sc_example)$Timepoint <- NA # initialize a new pData column

pData(ex_sc_example)[grep("0hr", rownames(pData(ex_sc_example))),"Timepoint"] <- "0hr"
pData(ex_sc_example)[grep("1hr", rownames(pData(ex_sc_example))),"Timepoint"] <- "1hr"
pData(ex_sc_example)[grep("4hr", rownames(pData(ex_sc_example))),"Timepoint"] <- "4hr"

plot_tsne_metadata(ex_sc_example, gene = "Tnf", title = "Tnf over Time", facet_by = "Timepoint", size = 0.2)
# Notice that now we are including a facet that will break the tSNE plot into smaller plots, one each for the variable. From this we can see that some clusters are forming based on timepoint, while others contain multiple timepoints.

plot_violin(ex_sc_example, title = "Tnf across clusters", gene = "Tnf", color_by = "Timepoint", facet_by = "Cluster", size = 1, ncol = 3)

ex_sc_example_subset <- subset_ex_sc(ex_sc_example, variable = "Cluster", select = c("Cluster4")) #subsetting can be very handy! If needed chain together several rounds of subsetting.

plot_tsne_metadata(ex_sc_example_subset, gene = "Ccl22", facet_by = "Timepoint", title = "Cluster 4 cells") # this function is flexible for plotting any data onto a 2d tsne mapping

plot_violin(ex_sc_example_subset, title = "Ccl22 across time in Cluster 4 cells", gene = "Ccl22", facet_by = "Cluster", size = 1, ncol = 3, color_by = "Timepoint")

```

## Identify markers and reference genes

For normalization and cluster id a function is provided that will id both marker genes and reference genes.

```{r, error=FALSE, warning=FALSE, cache=FALSE, include=TRUE}
ex_sc_example <- id_markers(ex_sc_example, num_markers = 10, num_reference = 1000) # This is a quick method to find good markers genes for cell identification. These genes get written to fData()

marker_list <- return_markers(ex_sc_example) # However returning the genes as a list can be helpful

plot_scatter(input = ex_sc_example, title = "Top 2 marker (Alphabetical) genes for Cluster 4", gene1 = marker_list$Cluster4[1], gene2 = marker_list$Cluster4[2], facet_by = "Cluster", color_by = "Cluster") # A scater plotting function. Note the correlation of the marker genes.

marker_facet <- c("Csf1r", "Lcn2", "Ccr7")

plot_tsne_multigene(input = ex_sc_example, genes = marker_facet, title = "Marker Genes",  ncol = 3)

```

