---
title: "SignallingSingleCell"
geometry: margin=2cm
author: "Kyle Gellatly"
date: "`r Sys.Date()`"
output:
  html_document:
    highlight: tango
    theme: cerulean
    toc: yes
    toc_float : yes
# pdf_document:
#   highlight: tango
#   toc: yes
# output: rmarkdown::html_vignette
# vignette: >
#   %\VignetteIndexEntry{Vignette Title}
#   %\VignetteEngine{knitr::rmarkdown}
#   %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE, warning=FALSE, error=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "")
```


```{r, include=FALSE, cache=FALSE, warning=FALSE, error=FALSE}
library("SignallingSingleCell")
load("~/Documents/Work/Garber/Data/inDrop/mDC_UMIClean/1-RawFiles/mDC_0hr_1hr_4hr_CLEAN.Rdata")
sc_dat <- mDC_0hr_1hr_4hr_CLEAN
```

# Preprocessing

## Filtering, OPTIONAL

The first step is to filter your data to remove low quality cells. Often creating a histogram of the values and assigning cutoffs is simple and effective. Typically we remove all cells lower than 1000 UMIs / cell.

```{r, include=TRUE, cache=FALSE, warning=FALSE, error=FALSE}
dim(sc_dat)
UMI_sum <- apply(sc_dat,2,sum)
range(UMI_sum)
qplot(UMI_sum, geom = "histogram")  

filtered_data <- pre_filter(input = sc_dat, minUMI = 1000, maxUMI = 7500, threshold = 1, minCells = 10)  

dim(filtered_data)
UMI_sum <- apply(filtered_data,2,sum)
range(UMI_sum)
qplot(UMI_sum, geom = "histogram")
```

## Constructing the ExpressionSet Class, MANDATORY  

The ExpressionSet class (ex_sc) is an extremely convienient data structure that contains 3 dataframes. Now that we have a filtered dataset it can be stored in this format.  

exprs(ex_sc) is the expression data  
pData(ex_sc) is cell information, where rows are cells and columns are metadata  
fData(ex_sc) is gene information, where rows are genes and columns are metadata  
ncol(exprs(ex_sc)) == nrow(pData(ex_sc))  
nrow(exprs(ex_sc)) == nrow(fData(ex_sc))  

Now that we have our first information (UMI counts per cell) we can write that into the pData.

```{r, include=TRUE, cache=FALSE, warning=FALSE, error=FALSE}
ex_sc_example <- construct_ex_sc(filtered_data)  

pData(ex_sc_example) <- cbind(pData(ex_sc_example), UMI_sum)
head(pData(ex_sc_example))
```

## Initial Dimension reduction, MANDATORY

Before normalization preliminary dimensionality reduction is necessary to form preliminary clusters. These clusters are used to normalize internal to a cluster before normalizing across clusters. First we can subset the genes, and then use these feature selected genes for dimension reduction.

```{r, include=TRUE, cache=FALSE, warning=FALSE, error=FALSE}
gene_subset <- subset_genes(input = exprs(ex_sc_example), method = "Expression", threshold = 5, minCells = 50)  

prelim_dims <- dim_reduce(input = exprs(ex_sc_example)[gene_subset,], pre_reduce = "iPCA", nComp = 15, tSNE_perp = 30, print_progress=FALSE)  

pData(ex_sc_example) <- cbind(pData(ex_sc_example), prelim_dims)
head(pData(ex_sc_example))
```

From this we can map on our UMI counts before normalization to view if there is any bias. We can also look at a gene of interest. To accomplish these we will take advantage of the ex_sc set once again.

```{r, include=TRUE, cache=FALSE, warning=FALSE, error=FALSE}
prelim_dims_tsne <- pData(ex_sc_example)[,c("x", "y", "UMI_sum")]
ggplot(prelim_dims_tsne) + 
  geom_point(aes(x = x, y = y, col = UMI_sum)) +
  scale_color_gradientn(colours=c("gray", 'blue', 'red', 'yellow')) +
  theme_classic() + 
  labs(title= "Initial Positions", x = "tSNE[1]", y = "tSNE[2]")  

gene <- "Lcn2"
prelim_dims_tsne$gene <- log2(filtered_data[gene,]+2)-1
ggplot(prelim_dims_tsne) + 
  geom_point(aes(x = x, y = y, col = gene)) +
  scale_color_gradientn(colours=c("gray", 'blue', 'red', 'yellow')) +
  theme_classic() + 
  labs(title= gene, x = "tSNE[1]", y = "tSNE[2]")
```

After viewing a large UMI bias we can now use spectral clustering for intial clustering before normalization. It looks as though there are about 6 clusters.

## Initial Clustering, MANDATORY

```{r, include=TRUE, cache=FALSE, warning=FALSE, error=FALSE}
prelim_dims_PCA <- as.data.frame(pData(ex_sc_example)[,grep("PC", colnames(pData(ex_sc_example)))])
clusters <- cluster_sc(input = prelim_dims_PCA, method = "spectral", num_clust = 6)  

pData(ex_sc_example)$Cluster <- paste0("Cluster", clusters)  

ggplot(pData(ex_sc_example)) + 
  geom_point(aes(x = x, y = y, col = Cluster)) +
  theme_classic() + 
  labs(title= "Initial Clusters", x = "tSNE[1]", y = "tSNE[2]")
```
