---
title: "SignallingSingleCell"
geometry: margin=2cm
author: "Kyle Gellatly"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "")
```


```{r, include=FALSE, cache=FALSE}
library("SignallingSingleCell")
load("~/Documents/Work/Garber/Data/inDrop/mDC_UMIClean/1-RawFiles/mDC_0hr_1hr_4hr_CLEAN.Rdata")
sc_dat <- mDC_0hr_1hr_4hr_CLEAN
```

# Preprocessing

## Filtering

The first step is to filter your data to remove low quality cells. Often creating a histogram of the values and assigning cutoffs is simple and effective. Typically we remove all cells lower than 1000 UMIs / cell.

```{r, include=TRUE, cache=FALSE, warning=FALSE, error=FALSE}
dim(sc_dat)
UMI_sum <- apply(sc_dat,2,sum)
range(UMI_sum)
qplot(UMI_sum, geom = "histogram")

filtered_data <- pre_filter(input = sc_dat, minUMI = 1000, maxUMI = 7500, threshold = 1, minCells = 10)

dim(filtered_data)
UMI_sum <- apply(filtered_data,2,sum)
range(UMI_sum)
qplot(UMI_sum, geom = "histogram")
```

## Constructing the ExpressionSet Class

The ExpressionSet class (ex_sc) is an extremely convienient data structure that contains 3 dataframes. Now that we have a filtered dataset it can be stored in this format.

exprs(ex_sc) is the expression data
pData(ex_sc) is cell information, where rows are cells and columns are metadata
fData(ex_sc) is gene information, where rows are genes and columns are metadata
ncol(exprs(ex_sc)) == nrow(pData(ex_sc))
nrow(exprs(ex_sc)) == nrow(fData(ex_sc))

Now that we have our first information (UMI counts per cell) we can write that into the pData.

```{r, include=FALSE, cache=FALSE}
ex_sc_example <- construct_ex_sc(filtered_data)
pData(ex_sc_example) <- cbind(pData(ex_sc_example), UMI_sum)
head(pData(ex_sc_example))
```

## Initial Dimension reduction

Before normalization preliminary dimensionality reduction is necessary to form preliminary clusters. These clusters are used to normalize internal to a cluster before normalizing across clusters. First we can subset the genes, and then use these feature selected genes for dimension reduction.

```{r, include=TRUE, cache=FALSE, warning=FALSE, error=FALSE}
gene_subset <- subset_genes(exprs(ex_sc_example), threshold = 5, minCells = 50)
prelim_dims <- dim_reduce(input = exprs(ex_sc_example)[gene_subset,], pre_reduce = "ICA", nComp = 10, tSNE_perp = 30, print_progress=TRUE)
pData(ex_sc_example) <- cbind(pData(ex_sc_example), prelim_dims)
head(pData(ex_sc_example))
```

From this we can map on our UMI counts before normalization to view if there is any bias. We can also look at a gene of interest. To accomplish these we will take advantage of the ex_sc set once again.

```{r, include=TRUE, cache=FALSE, warning=FALSE, error=FALSE}
prelim_dims_tsne <- pData(ex_sc_example)[,c("x", "y")]
prelim_dims_tsne$UMI_sum <- pData(ex_sc_example)$UMI_sum
ggplot(prelim_dims_tsne) + 
  geom_point(aes(x = x, y = y, col = UMI_sum)) +
  scale_color_gradientn(colours=c("gray", 'blue', 'red', 'yellow')) +
  theme_classic() + 
  labs(title= "Initial Positions", x = "tSNE[1]", y = "tSNE[2]")

gene <- "Lcn2"
prelim_dims_tsne$gene <- log2(filtered_data[gene,]+2)-1
prelim_dims_tsne <- prelim_dims_tsne[with(prelim_dims_tsne, order(prelim_dims_tsne[,"gene"])), ]

ggplot(prelim_dims_tsne) + 
  geom_point(aes(x = x, y = y, col = gene)) +
  scale_color_gradientn(colours=c("gray", 'blue', 'red', 'yellow')) +
  theme_classic() + 
  labs(title= gene, x = "tSNE[1]", y = "tSNE[2]")
```

After viewing a large UMI bias we can now use spectral clustering for intial clustering before normalization. It looks as though there are about 5 clusters.

## Initial Spectral Clustering


```{r, include=TRUE, cache=FALSE, warning=FALSE, error=FALSE}
prelim_dims_ICA <- as.data.frame(prelim_dims[,grep("IC", colnames(pData(ex_sc_example)))])
spec <- kknn::specClust(prelim_dims_ICA, centers = 5, method = 'random-walk')
prelim_dims_tsne$Cluster <- paste0("Cluster", spec$cluster)

ggplot(prelim_dims_tsne) + 
  geom_point(aes(x = x, y = y, col = Cluster)) +
  theme_classic() + 
  labs(title= "Initial Clusters", x = "tSNE[1]", y = "tSNE[2]")
```
